### 2019-02-12 sections/comparison.tex
compact code examples
first stab at integrating processes in intuition section

This section gives an overview of the essential concepts of \TOP and how they are represented in \TOPHAT.
While \TOP is focussed on collaboration, it gives rise to concepts like communication and multiprogramming.
% These concepts are: communication with the environment, communication between subsystems, sequential composition, concurrency, and synchronisation.
These concepts can also be found in process algebras, which raises the question how \TOPHAT and process algebras relate.
While introducing the concepts, we compare them with Hoare's Communicating Sequential Processes (\CSP).
The comparison is based on Hoare's book \cite{books/Hoare85CSP}.
We chose \CSP as an example to stand for the numerous process algebras in existence, which, as far as this section is concerned, are reasonably similar.

We provide comparisons based on two aspects.
First, we compare the scope of the languages, as intended by the authors.
Second, we focus on features common to \TOP and \CSP, and study their realisation.
When we point out differences, we do not argue that the different realisations are incompatible.
As a matter of fact the primitives can certainly be expressed in terms of each other.
Instead, we point out how the systems emphasise certain points of view by choosing different building blocks to express the same concepts.

\paragraph{Scope}

The central objective of \TOP is to coordinate collaboration.
Therefore it has combinators to express ways in which people can work together to reach a common goal.
Tasks can be executed after each other, at the same time, or conditionally.
This motivates the task combinators for steps, parallel combination, and choice.
To employ these patterns of collaboration, users need to communicate.
This motivates the different forms of communication: with the environment, along control flow, and across control flow.
\TOPHAT is a language to specify workflows.

The central objective of \CSP is to model the patterns of behaviour of processes.
These patterns of behaviour manifest themselves in sequences of actions that processes can engage in.
\CSP is a language to specify synchronisation patterns.

\begin{quote}
\noindent In \TOP, communication is a means to an end. \\
In \CSP, communication is the central matter of discourse.
\end{quote}

%%NOTE: Focus on what TOP is good at --TS
% \CSP has a formal semantics that allows various kinds of correctness proofs, including equality of processes, and adherence to a specification.
% This allows applications in program correctness, proofs of deadlock freedom, liveness, or verification of protocols.
%
% \TOP focuses less on formal correctness, and more on practical applicability.
The big difference between \CSP and \TOP is the intended level of abstraction.
While \CSP is used to explicitly model low-level details about communication, \TOPHAT wants to hide details and provide high-level communication mechanisms to the programmer.
It wants to be a calculus with intuitive semantics that facilitates communication between programmers and domain experts.
Such programs are supposed to hide implementation details from domain experts while containing enough information to allow automatic generation of executable applications, including user interfaces.

@@

Both \TOP and \CSP have two sorts of communication: with the environment, and between subsystems.
In both languages, communication with the \emph{environment} is input and output.
They block evaluation of programs until the environment sends or receives events.
In both languages, communication \emph{between subsystems} does not block.
They evaluate programs as far as possible until further external input is required.

@@

In \TOPHAT, communication with the environment is represented by editors.
A task of the form $\Edit v$, where $v$ is of type $\tau$, can receive any change event with a value of type $\tau$ and additionally the special event $\Empty$ that empties the editor.
The environment can always inspect the current value of an editor.

@@

In \CSP, communication with the environment is represented by prefixing.
The process $(a \to P)$ can engage in action $a$, after which it continues as process $P$.
%%NOTE: Not that important I think, we don't use it in the rest of the text. --TS
% In general, if $B$ is a set of events and $P(x)$ is an expression that evaluates to a process for each $x \in B$, the process $(x:B \to P(x))$ can engage in any one of the actions in $B$, after which it continues as the process determined by $P(x)$.
% Hoare does not specify the language in which $P(x)$ is to be expressed.
% He seems to permit any kind of mathematical formula, or any programming language.
%
Prefixing has no direction of sending and receiving.
Hoare uses the neutral phrase \emph{engaging in an action}.
The interpretation of whether an action stands for input or output is left to the reader.
For example, a vending machine $P = (\textit{coin} \to \textit{choc} \to P)$ is to be interpreted as taking a coin as input and producing chocolate as output.

Sending and receiving of values is modelled by giving additional structure to the names of actions.
For example, an action name could be $\textit{in}.5$, which can be interpreted as inputting the value 5.
%
In \TOP, event values are typed to match the type of the receiving editor.
In \CSP, action names are essentially strings, subject to implicit restrictions.

@@

This example demonstrates external communication and choice.
The following \CSP program models a vending machine that dispenses a small biscuit for one coin and a large biscuit for two coins.
\begin{equation*}
  V = \textit{coin} \to (\textit{small} \to \text{STOP} \mid \textit{coin} \to \textit{large} \to \text{STOP})
\end{equation*}

A vending machine in \TOPHAT that is close in spirit to the one above looks as follows.
\begin{TASK}
  enter Int >>? \n. if n == 1 then edit "small"
    else if n == 2 then edit "large" else fail
\end{TASK}
This machine does not exhibit exactly the same behaviour as the one in \CSP, nor is it intended to.
Instead, it just demonstrates how the same choice
%%NOTE: Problem is already stated --TS
% of buying a small biscuit for one coin or a large biscuit for two coins
can be expressed in \TOP.
The task asks the user to enter an amount of money.
This editor stands for a coin slot in a real machine that freely accepts and returns coins.
There is a continue button that is initially disabled.
When the user has inserted exactly 1 or 2 coins, the continue button becomes enabled.
When the user presses the continue button, the machine dispenses either a large or a small biscuit, depending on the amount of money.

@@

\TOP has two methods for internal communication, both different from its meth\-od for external communication.
The step combinator represents data flow that follows control flow, and shared data represents data flow across control flow.

The step combinators in $t \Then e$ and $t \Next e$ pass the value of $t$ to the continuation $e$.
% The intuition about sequential composition is to first do one thing, and once it is done, do some other thing.
% In \TOP, tasks never terminate.
% Nonetheless, the notion of sequential composition exists in two forms: the external and the internal step.
They both act like $t$ as long as the step is guarded.
Once it becomes unguarded, the task continues as $e\ v$, where $v$ is the value of $t$.
The task $t \Next e$ requires an input event $\Continue$ in addition to the step being unguarded. % in order to step.
% \CSP's semicolon is like an internal step $t \Then e$ in \TOP where $t$ uses $\Unit$ to signal when it is done, and $e$ never fails.

Shared data sources are assignable references, whose changes in value are immediately visible to all tasks interested in them.
The semantics of \TOPHAT requires all updates to shared data and all enabled internal steps to be processed before any further communication with the environment can take place.
This introduces the possibility of diverging computations, where a cyclic dependency between shared data causes internal steps to be taken continuously without ever being able to engage in external communication.

\CSP uses prefixing for both internal and external communication.
Prefixing becomes internal by using \emph{concealment}.
Processes must take any concealed steps as soon as possible, before any further communication with the environment can take place.
This means concealed actions have priority over exposed ones.
This introduces the possibility of diverging computations, that is processes which continuously take internal steps without ever being able to engage in external communication.

@@

\subsection{Multiprogramming}

Both \TOP and \CSP are models of \emph{multiprogramming}, as opposed to \emph{multiprocessing}.
This means that one processor evaluates multiple programs in an interleaving fashion.
% instead of multiple processors.
Multiprogramming involves three aspects: concurrency, synchronisation, and nondeterminism.

\paragraph{Concurrency}

Concurrency is the source of parallelism: two things are done at the same time.
%
In \TOP, the parallel combinator $t_1 \And t_2$ stands for two independent tasks carried out at the same time.
For the system it does not matter if the tasks are executed by two people actually in parallel, or by one person who switches between the tasks.
All events sent to the combined task are interleaved into a serial stream of inputs.
The tasks are truly independent of each other, all interleavings are permitted.
Events to $t_1$ and $t_2$ must be prefixed by $\First$(irst) and $\Second$(econd) respectively.
This unambiguously renames the events, removing any source of nondeterminism.

@@

\paragraph{Synchronisation}

@@

\paragraph{Nondeterminism}

@@

\paragraph{Conclusion}

>>>>

%NOTE: Use \emph{process} as a general term for a process in process algebras and a reactive system from reactive programming?
%NOTE: Main focus should be on \emph{collaboration} between tasks opposed to \emph{communication} between processes.

This section gives an overview of the essential concepts of tasks and how they are represented in \TOPHAT.
While tasks are focussed on collaboration, they give rise to concepts like communication, concurrency and synchronisation.
% These concepts are: communication with the environment, communication between subsystems, sequential composition, concurrency, and synchronisation.
These concepts can also be found in process algebras like \CSP, \CCS and \PICALC, \fixme{cite}
and reactive programming languages like \ESTEREL, \HIPHOP, and \fixme{more?}.
This raises the question how \TOPHAT relates to these algebras and languages.

In the next subsections, we give an intuition about tasks, by discussing the concepts collaboration, communication \fixme{and multiprocessing}.
While introducing these concepts, we compare them with \emph{processes}.
We define a process as \fixme{definition}.
% We chose processes as an example to stand for the numerous process algebras in existence, which, as far as this section is concerned, are reasonably similar.
We focus on features common to tasks and processes, and study their realisation.
When we point out differences, we do not argue that the different realisations are incompatible.
As a matter of fact, primitives can certainly be expressed in terms of each other.
Instead, we point out how the systems emphasise certain points of view by choosing different building blocks to express the same concepts.

% we provide comparisons based on two aspects.
% First, we compare the scope of the languages, as intended by the authors.

\subsection{Collaboration}

The central objective of tasks is to coordinate collaboration.
Therefore it has combinators to express ways in which people can work together to reach a common goal.
Tasks can be executed after each other, at the same time, or conditionally.
This motivates the task combinators for steps, parallel combination, and choice.
To employ these patterns of collaboration, users need to communicate.
This motivates the different forms of communication: with the environment, along control flow, and across control flow.
Tasks are a way to model workflows.
% \TOPHAT is a language to specify workflows.

The central objective of processes is to model patterns of their behaviour.
These patterns of behaviour manifest themselves in sequences of actions that processes can engage in.
Actions can be events from the environment, or signals send internally between processes.
Processes are a way to model synchronisation patterns.

% \begin{quote}
%   \noindent In \TOP, communication is a means to an end. \\
%   In \CSP, communication is the central matter of discourse.
% \end{quote}

%%NOTE: Focus on what TOP is good at --TS
% \CSP has a formal semantics that allows various kinds of correctness proofs, including equality of processes, and adherence to a specification.
% This allows applications in program correctness, proofs of deadlock freedom, liveness, or verification of protocols.
%
% \TOP focuses less on formal correctness, and more on practical applicability.
The big difference between processes and tasks is the intended level of abstraction.
While processes are used to explicitly model low-level details about communication, tasks hide these details and provide high-level communication mechanisms to the programmer.
It wants to be a calculus with intuitive semantics that facilitates communication between programmers and domain experts.
Such programs are supposed to hide implementation details from domain experts while containing enough information to allow automatic generation of executable applications, including user interfaces.

@@

Both tasks and processes have two sorts of communication: with the environment, and between subsystems.
In both paradigms, communication with the \emph{environment} is input and output.
They block evaluation of programs until the environment sends or receives events.
In both paradigms, communication \emph{between subsystems} does not block.
They evaluate programs as far as possible until further external input is required.

@@

In \TOPHAT, communication with the environment is represented by editors.
A task of the form $\Edit v$, where $v$ is of type $\tau$, can receive any change event with a value of type $\tau$ and additionally the special event $\Empty$ that empties the editor.
There is no explicit communication of output,
However, the environment can always inspect the current value of an editor.

@@

Processes communicate with the environment by sending or receiving actions
The \CSP process $(a \to P)$ can engage in action $a$, after which it continues as process $P$.
% Prefixing has no direction of sending and receiving.
Hoare uses the neutral phrase \emph{engaging in an action}.
The interpretation of whether an action stands for input or output is left to the reader.
For example, a vending machine $P = (\textit{coin} \to \textit{choc} \to P)$ is to be interpreted as taking a coin as input and producing chocolate as output.
%%NOTE: Not that important I think, we don't use it in the rest of the text. --TS
% In general, if $B$ is a set of events and $P(x)$ is an expression that evaluates to a process for each $x \in B$, the process $(x:B \to P(x))$ can engage in any one of the actions in $B$, after which it continues as the process determined by $P(x)$.
% Hoare does not specify the language in which $P(x)$ is to be expressed.
% He seems to permit any kind of mathematical formula, or any programming language.

The equivalent process in \ESTEREL is the reactive statement $\text{await} a; P$.
Here, action $a$ is interpreted as an input.
To emit an action, one can use $\text{emit} a$.

Sending and receiving of values is modelled by giving additional structure to the names of actions.
For example, an action name could be $\textit{in}.5$, which can be interpreted as inputting the value 5.
%
In \TOP, input values are typed to match the type of the receiving editor.
In \CSP and \ESTEREL, action names are essentially strings, subject to implicit restrictions.

@@

This example demonstrates external communication and choice.
The following \CSP program models a vending machine that dispenses a small biscuit for one coin and a large biscuit for two coins.
\begin{equation*}
  V_\text{\CSP} = \textit{coin} \to (\textit{small} \to \text{STOP} \mid \textit{coin} \to \textit{large} \to \text{STOP})
\end{equation*}
An equivalent reactive system in \ESTEREL could be defined as\fixme{validate}
\begin{equation*}
  V_\text{\ESTEREL} = \text{await } \textit{coin}; ( \text{await } \textit{button}; \text{emit } \textit{small} \parallel \text{await } \textit{coin}; \text{await } \textit{button}; \text{emit } \textit{large})
\end{equation*}

A vending machine in \TOPHAT that is close in spirit to the one above looks as follows.
\begin{TASK}
  enter Int >>? \n. if n == 1 then edit "small" else if n == 2 then edit "large" else fail
\end{TASK}
This machine does not exhibit exactly the same behaviour as the ones in \CSP or \ESTEREL, nor is it intended to.
Instead, it just demonstrates how the same choice
%%NOTE: Problem is already stated --TS
% of buying a small biscuit for one coin or a large biscuit for two coins
can be expressed in \TOPHAT.
The task asks the user to enter an amount of money.
This editor stands for a coin slot in a real machine that freely accepts and returns coins.
There is a continue button that is initially disabled.
When the user has inserted exactly 1 or 2 coins, the continue button becomes enabled.
When the user presses the continue button, the machine dispenses either a large or a small biscuit, depending on the amount of money.

@@

\TOPHAT has two methods for internal communication, both different from its method for external communication.
The step combinator represents data flow that follows control flow, and shared data represents data flow across control flow.
%
The step combinators in $t \Then e$ and $t \Next e$ pass the value of $t$ to the continuation $e$.
% The intuition about sequential composition is to first do one thing, and once it is done, do some other thing.
% In \TOP, tasks never terminate.
% Nonetheless, the notion of sequential composition exists in two forms: the external and the internal step.
They both act like $t$ as long as the step is guarded.
Once it becomes unguarded, the task continues as $e\ v$, where $v$ is the value of $t$.
The task $t \Next e$ requires an input event $\Continue$ in addition to the step being unguarded. % in order to step.
% \CSP's semicolon is like an internal step $t \Then e$ in \TOP where $t$ uses $\Unit$ to signal when it is done, and $e$ never fails.

\fixme{Add short intro on communication between non-sequential tasks.}
Shared data sources are assignable references, whose changes in value are immediately visible to all tasks interested in them.
The semantics of \TOPHAT requires all updates to shared data and all enabled internal steps to be processed before any further communication with the environment can take place.
This introduces the possibility of diverging computations, where a cyclic dependency between shared data causes internal steps to be taken continuously without ever being able to engage in external communication.

\fixme{What to do with this paragraph?}
\CSP uses prefixing for both internal and external communication.
Prefixing becomes internal by using \emph{concealment}.
Processes must take any concealed steps as soon as possible, before any further communication with the environment can take place.
This means concealed actions have priority over exposed ones.
This introduces the possibility of diverging computations, that is processes which continuously take internal steps without ever being able to engage in external communication.

@@

% \subsection{Multiprogramming}

\subsection{Concurrency}

Both tasks and processes are models of \emph{multiprogramming}, as opposed to \emph{multiprocessing}.
This means that one processor evaluates multiple programs in an interleaving fashion.
% instead of multiple processors.
Multiprogramming involves three aspects: concurrency, synchronisation, and nondeterminism.

Concurrency is the source of parallelism: two things are done at the same time.
%
Using tasks $t_1$ and $t_2$, the parallel combination $t_1 \And t_2$ stands for two independent tasks carried out at the same time.
For the system it does not matter if the tasks are executed by two people actually in parallel, or by one person who switches between the tasks.
All events sent to the combined task are interleaved into a serial stream of inputs.
The tasks are truly independent of each other, all interleavings are permitted.
Events to $t_1$ and $t_2$ must be prefixed by $\First$(irst) and $\Second$(econd) respectively.
This unambiguously renames the events, removing any source of nondeterminism.

@@

\subsection{Synchronisation}

@@

\subsection{Nondeterminism}

@@

\subsection{Recap}
